---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# avonet

<!-- badges: start -->
<!-- badges: end -->

[AVONET](https://figshare.com/s/b990722d72a26b5bfead) is a morphological, ecological and geographical dataset for all birds. For more information on how the dataset was compiled, please refer to [Tobias, et al. 2022](https://onlinelibrary.wiley.com/doi/10.1111/ele.13898).

The goal of the `avonet` package is to give users access to AVONET data comfortably from within an R session. It queries a Postgres database directly, so there are no flat files to download or keep in sync: you connect once, then pull traits for any taxon, filter them, and get the literature sources needed to cite them back alongside the data.

## Installation

You can install the development version of `avonet` from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("kfrac/avonet")
```

## Connecting

Connecting to the AVONET database is as easy as entering your username and password. We recommend using the `keyring` package to manage your credentials.

```{r DB_connection}
library(avonet)
library(keyring)
db_user <- key_list("avonet")[1, "username"]
db_password <- key_get("avonet", username = db_user)
connect_db(username = db_user, pw = db_password)
```

Every function below assumes an open connection. Bird taxonomy is not settled, so AVONET carries three parallel taxonomies and every query takes one: BirdLife (`1`), eBird (`2`) or BirdTree (`3`).

## Finding out what is available

Traits are organised into groups, and each group can be listed with its descriptions, possible values and primary source.

```{r trait_groups}
get_trait_groups()
```

```{r trait_list}
get_trait_list(group = "eco")
```

For a categorical trait, the valid levels can be pulled directly — useful for building a filter without guessing at spelling.

```{r trait_levels}
get_trait_levels("habitat")
```

Passing `return_defs = TRUE` returns what each level actually means, rather than just its name.

```{r trait_level_defs}
get_trait_levels("trophic_niche", return_defs = TRUE)
```

## Querying species-level data

`get_species()` is the lightweight path: give it species binomials and it returns one row per species.

```{r example_single_query}
get_species("Buteo buteo", 1)
```

Multiple species can also be queried simultaneously.

```{r example_multi_query}
my_birds <- c("Haliaeetus leucocephalus", "Aptenodytes forsteri", "Cardinalis cardinalis")
get_species(my_birds, 1)
```

Each trait in the database carries a source. These are stripped by default, since they hold source IDs rather than values, but `source_cols = TRUE` keeps them inline.

```{r example_source_cols}
names(get_species("Buteo buteo", 1, source_cols = TRUE))
```

## Querying any taxonomic rank

`get_traits()` is the main entry point. Unlike `get_species()` it accepts names at any rank — species, genus, family or order — and works out which is which, so a single call can mix them.

```{r traits_basic}
buteo <- get_traits("Buteo buteo", taxonomy = 1)
names(buteo)
```

The result is a list. `data` holds one row per species, `metadata_summary` describes each trait and where it came from, and `detailed_sources` holds the full literature references behind those sources — so a result can be cited without a second lookup.

```{r traits_data}
buteo$data[, c("species", "family", "habitat", "trophic_niche", "mass_value")]
```

```{r traits_metadata}
head(buteo$metadata_summary, 4)
```

Ranks are detected per name, and reported so you can see how an ambiguous name was read.

```{r traits_rank}
hawks <- get_traits("Accipitridae", taxonomy = 1)
nrow(hawks$data)
```

## Filtering

Filters are a named list, applied to the data after it returns. Categorical traits take a value or a set of values; numeric traits take an operator and a value. Conditions combine with AND.

```{r traits_filter, message = FALSE}
forest_hawks <- get_traits(
  "Accipitridae",
  taxonomy = 1,
  filter = list(habitat    = "Forest",
                range_size = list(op = "<", val = 5000000))
)

head(forest_hawks$data[, c("species", "habitat", "range_size", "mass_value")])
```

Trait names can be given in short form (`habitat`, `range_size`, `mass`) or as their full database column names. Filtering on a column that was not queried is an error rather than an empty result, and the message lists what is available.

## Specimen-level measurements

Ecological, mass and geographic traits are recorded once per species. Morphological measurements are recorded per measured individual, so there are many rows per species. Rather than repeating every species-level value across each specimen, `resolution = "specimen"` returns the measurements as a separate element and leaves `data` at one row per species.

```{r traits_specimen, message = FALSE}
specimens <- get_traits("Buteo buteo", taxonomy = 1, resolution = "specimen")
names(specimens)
head(specimens$specimen_data)
```

Supplying `aggregate` collapses those measurements into per-species means and counts for a given grouping — `"sex"`, `"life stage"`, `"country"`, `"source type"` or `"all"`.

```{r traits_aggregate, message = FALSE}
by_sex <- get_traits(
  "Buteo buteo",
  taxonomy = 1,
  resolution = "specimen",
  aggregate  = "sex"
)

by_sex$specimen_data
```

## Checking a name before querying

Taxonomies disagree, and names get split, merged and renamed. `get_taxonomic_info()` shows what a search term resolves to, and how it matched.

```{r taxonomic_info}
get_taxonomic_info(search_term = "Buteo buteo", taxonomy = 1)
```

```{r taxonomic_info_rank}
head(get_taxonomic_info(search_term = "Buteo", taxonomy = 1))
```

## Disconnecting

Remember to disconnect from the database at the end!

```{r DB_disconnect}
disconnect_db()
```
