---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# avonet

<!-- badges: start -->
<!-- badges: end -->

[AVONET](https://figshare.com/s/b990722d72a26b5bfead) is a morphological, ecological and geographical dataset for all birds. For more information on how the dataset was compiled, please refer to [Tobias, et al. 2022](https://onlinelibrary.wiley.com/doi/10.1111/ele.13898).

The goal of the `avonet` package is to give users access to AVONET data comfortably from within an R session. For now, it provides functions for connecting to the database and querying data. Future work will include functions to filter and aggregate data on different taxonomic levels.

## Installation

You can install the development version of `avonet` from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("kfrac/avonet")
```

## Usage

Connecting to the AVONET database is as easy as entering your username and password.

```{r DB_connection}
library(avonet)
con <- connect(username = "postgres", pw = "Frankfurterstr25!")
```

For now, you can query bird species according to 3 different taxonomies: BirdLife (1), eBird (2) and BirdTree (3).

```{r example_single_query}
get_species(con, "Buteo buteo", 1)
```

Multiple species can also be queried simultaneously.

```{r example_multi_query}
my_birds <- c("Haliaeetus leucocephalus", "Aptenodytes forsteri", "Cardinalis cardinalis")
result_df <- get_species(con, my_birds, 1)
```

Remember to disconnect from the database at the end!

```{r DB_disconnect}
DBI::dbDisconnect(con)
```
